Data Specialization for a Staged Lambda Calculus

### 2014-11-06 paper/splitting.tex
Completed pass through 4.1

Although terms in \lang\ clearly distinguish their stage \bbone\ and stage
\bbtwo\ components, these components may be interleaved in \lang\ code. The
goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
of a term $e$ in \lang\ into a pair of
terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
stage \bbone\ subcomputations in $e$ and $e_2$ performs all of the stage \bbtwo\ subcomputations.
Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same value 
as evaluating $e$ in \lang.

Since the mechanics of splitting stage \bbtwo\ terms are conceptually simpler than those for stage \bbone\
terms, we begin by describing how to split stage \bbtwo\ terms before proceeding to more complicated cases.

@@

Splitting a stage \bbtwo\ term $\coltwo e A$ generates
 a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
of $e$ in place to create a residual, splitting statically isolates these computations into
$p$.  The resumer accepts the value produced by the precomputation as input
% (bound to the free variable $l$)
and executes all stage 2 subcomputations.   
Specifically, if $e$ in \lang\ reduces to the final value $v$, 
then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
(The resumer, bound to the output of $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})

@@

Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym
[p,l.r]$ (\cref{fig:termSplitTwo}), is guided by the structure of the term $e$.
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
\end{lstlisting}
has two subterms, the latter of which itself has two subterms. Two are $\pause$s
containing stage \bbone\ computations, while the third has no stage \bbone\
content.

\ur{The discussion of the subterms is not helpful. what is the point?}

The precomputation \emph{lifts} these computations in the same tree
structure: \verb|(1+2,(3+4,()))|.%
\ur{Not clear what a ``tree structure'' is. }
\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
The resumer is a function which computes the stage \bbtwo\ value of the original
term by mapping the results of the precomputation in place of the $\pause$s:
\verb|(L1,(L2,_)) => L1 < (L2 - 5)|.

\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}

In general, constants and variables have no stage \bbone\ content, so their
precomputations are \texttt{()}, and their resumers return the original term's
value. Non-staging constructs with $n$ subterms have precomputations which are
$n$-tuples of their subterms' precomputations, and resumers which pass each
component to the resumer of the respective subterm. (This behavior at
\texttt{case} and functions causes precomputations to be lifted out from
underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends on
splitting stage \bbone\ terms, and so they are covered later.

>>>>

Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
\bbtwo\ components, these components may be interleaved in \lang\ code. The
goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
of a term $e$ in \lang\ into a pair of
terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same value 
as evaluating $e$ in \lang.

Since the mechanics of splitting stage~\bbtwo\ terms are conceptually simpler than those for stage \bbone\
terms, we begin by describing how to split stage \bbtwo\ terms before proceeding to more complicated cases.

@@

Splitting a stage~\bbtwo\ term $\coltwo e A$ generates
 a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
of $e$ in place to create a residual, splitting statically isolates these computations into
$p$.  The resumer accepts the value produced by the precomputation as input
% (bound to the free variable $l$)
and executes all stage~\bbtwo\ subcomputations.   
Specifically, if $e$ in \lang\ reduces to the final value $v$, 
then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
(The resumer, bound to the output of $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})

@@

Splitting stage~\bbtwo\ terms, as implemented by the judgment $e \splittwosym
[p,l.r]$ in \cref{fig:termSplitTwo} is defined recursively on the structure of $e$.
In the terminal cases of constants and variables, splitting generates precomputations that are \texttt{()}
(these terms do not involve any stage~\bbone\ computations) and resumers that produce the original term's value.
For example, the integer constant \texttt{5} splits into the precomputation \texttt{()} and \texttt{l=>5}.

For all non-terminal stage~\bbtwo\ terms $e$ in \lang\ (except $\prev$ and $\pause$), splitting descends into $e$, recursively splitting subterms.  
The precompuation is formed by combining the resultant $n$ subterm precomputations into a tuple: $p=(p_1,\ldots,p_n)$.
The generated resumer has the same structure as $e$, with every subterm replaced by its resumer.
Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
underneath stage \bbtwo\ binders. % should probably draw a parallel to the same behavior in dynamics

The $\prev$ and $\pause$ rules depends on
splitting stage \bbone\ terms, and so they are covered later.

%For example, the stage \bbtwo\ term
%\begin{lstlisting}
%2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
%\end{lstlisting}
%has two subterms, the latter of which itself has two subterms. Two are $\pause$s
%containing stage \bbone\ computations, while the third has no stage \bbone\
%content. 
%
%\ur{The discussion of the subterms is not helpful. what is the point?}
%
%The precomputation \emph{lifts} these computations in the same tree
%structure: \verb|(1+2,(3+4,()))|.%
%\ur{Not clear what a ``tree structure'' is. }
%\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
%The resumer is a function which computes the stage \bbtwo\ value of the original
%term by mapping the results of the precomputation in place of the $\pause$s:
%\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
%
%\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
%
%Non-staging constructs with $n$ subterms have precomputations which are
%$n$-tuples of their subterms' precomputations, and resumers which pass each
%component to the resumer of the respective subterm.
